\bf Model Visualization Techniques for a Social Network Model

### 2017-04-12 paper.Rnw
add small ex

\section{Introduction} 
\st{
\begin{itemize}
%\item What? - model visualization and SAOMs. briefly dsecribe what these things are
%\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
\item How?
\end{itemize}
%also talk about other dynamic network models briefly to put SAOMs in context.
}

@@

\section{Networks and their Visualizations}

@@

There are two primary methods used to visualize networks. The first method, often referred to as node-link diagrams, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes. But, because there is no natural placement of the points, a random placement is used, then adjusted iteratively via a layout algorithm, of which there are many.

@@

The second primary method for network visualization uses the adjacency matrix of the network. This method is particularly useful when the network is very complex, dense, or large. Experimental studies have shown this to be true. For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs. This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, but a node-link diagram may underrepresent the edge count.

\section{Stochastic Actor-Oriented Models for Longitudinal Social Networks}

\st{What do I need to communicate in this section? 
\begin{itemize}
\item What the models are for: dynamic social networks with actor-level covariates
\item What are the model components? rate function, objective function (network statistics, actor covariates, interactions between covariates and network statistics), time points ($t_m = 1, \dots, M$)
\item What is RSiena? What methods of estimation, convergence checks does it use? What are the algorithms? What kinds of models can it fit? 
\end{itemize}}

@@

\section{Discussion}

>>>>

\section{Introduction} 
% \st{
% \begin{itemize}
% %\item What? - model visualization and SAOMs. briefly dsecribe what these things are
% %\item Why? - bring light to the RSiena blackbox. Need ways to visualize network models, collections of networks. 
% \item How?
% \end{itemize}
% %also talk about other dynamic network models briefly to put SAOMs in context.
% }

@@

In Section~\ref{sec:netintro}, we introduce the concepts of networks and network visualizations. In Section~\ref{sec:saoms}, we present the SAOM for social network analysis. Then, in Section~\ref{sec:ModelVis}, we combine concepts from Sections~\ref{sec:netintro} and \ref{sec:saoms} in an application of the model-vis paradigm, and conclude with a discussion in Section~\ref{sec:discussion}.

\section{Networks and their Visualizations} \label{sec:netintro}

@@

There are two primary methods used to visualize networks: node-link diagrams and matrix visualization. We will use the toy data set show below to demonstrate these two methods.

<<toyex>>=
set.seed(81393)
nodes <- data.frame(id = 1:5, group = sample(LETTERS[1:2], size = 5, replace = T), stringsAsFactors = F)
g <- igraph::as_data_frame(igraph::random.graph.game(n = 5, p = .4))
toynet <- list(nodes = nodes, edges = rbind(g, c(5,4)))
toynet
@

The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes. But, because there is no natural placement of the points, a random placement is used, then adjusted iteratively via a layout algorithm, of which there are many. The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.

@@

The second primary method for network visualization uses the adjacency matrix of the network. An example of this type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example. This method is particularly useful when the network is very complex, dense, or large. Experimental studies have shown this to be true. For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs. This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, but a node-link diagram may underrepresent the edge count with overlapping edges.

<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our undirected toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) + 
  geom_net(aes(from_id=from_id, to_id=to_id, shape =group, color = group), labelon = T, size = 7, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, arrowgap = .02) +
  scale_color_manual(values = c("red", "blue")) + 
  theme_net() + 
  labs(title = "Node-Link Diagram")

ggplot(data = toynet$edges) +
  geom_tile(aes(x = from, y = to), color = 'black', fill = 'black') +
  geom_tile(aes(x = to, y = from), color = 'black', fill = 'black') +
  geom_vline(xintercept = 0:5+.5) + 
  geom_hline(yintercept = 0:5+.5) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
  scale_y_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
  theme_bw() + 
  labs(title = "Adjacency Matrix Visualization") + 
  theme(aspect.ratio = 1)
@

\section{Stochastic Actor-Oriented Models for Longitudinal Social Networks} \label{sec:saoms}

% \st{What do I need to communicate in this section? 
% \begin{itemize}
% \item What the models are for: dynamic social networks with actor-level covariates
% \item What are the model components? rate function, objective function (network statistics, actor covariates, interactions between covariates and network statistics), time points ($t_m = 1, \dots, M$)
% \item What is RSiena? What methods of estimation, convergence checks does it use? What are the algorithms? What kinds of models can it fit? 
% \end{itemize}}

@@

\section{Discussion} \label{sec:discussion}
